---
title: "Getting Started with AnomalyPheno"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with AnomalyPheno}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE
)
```

## Overview

**AnomalyPheno** helps you understand *what kind* of anomaly occurred in a
time series, not just *whether* one occurred.

The pipeline has two stages:

1. **Detection.** Find anomalous intervals using a Spectral Residual or Local
   Outlier Factor engine.
2. **Phenotyping.** Group anomalies into labelled types using the
   Multi-Resolution Mapper algorithm, with confidence scores per assignment.

---

## Quick Start

```{r quickstart}
library(AnomalyPheno)

data(anomaly_benchmark)
result <- pheno_diagnose(anomaly_benchmark$x, q = 2.5)
print(result)
```

The returned `AnomPhenoResult` object contains:

| Field | Description |
|---|---|
| `$phenotypes` | One row per phenotype |
| `$instance_table` | One row per anomaly with confidence and stability |
| `$feature_importance` | 27 features ranked by F-statistic |
| `$canonical_templates` | Medoid segment per phenotype |

---

## Visualisations

```{r timeline}
plot(result, type = "timeline")
plot(result, type = "gallery")
plot(result, type = "profile")
plot(result, type = "landscape")
```

---

## Detection Engines

Two engines are available via the `engine` argument to `pheno_diagnose()`:

| Engine | Best for | Threshold `q` |
|---|---|---|
| `"sr"` (default) | Sharp spikes | q-sigma multiplier (2-4) |
| `"lof"` | Gradual drifts | LOF score cutoff (1.5-3) |

```{r engines}
res_lof <- pheno_diagnose(anomaly_benchmark$x, q = 2.0, engine = "lof")
```

---

## Prospective Classification

Once phenotypes are established, classify new anomaly segments without
re-running the full pipeline:

```{r predict}
pred <- pheno_predict(result, new_segment = c(0.1, 7.5, 0.2))
cat("Phenotype:", pred$phenotype_name, "| Confidence:", pred$confidence)
```

---

## Tidy Export

```{r tidy}
df <- as.data.frame(result)
head(df[df$is_anomaly, ])
```

---

## External Validation

When ground-truth labels are available:

```{r evaluate}
ev <- pheno_evaluate(result, anomaly_benchmark$true_labels)
print(ev)
```

---

## Synthetic Data

Generate controlled benchmarks with known phenotypes:

```{r simulate}
sim   <- pheno_simulate(n = 400, seed = 42)
res_s <- pheno_diagnose(sim$x, q = 2.5)
ev_s  <- pheno_evaluate(res_s, sim$true_labels)
cat(sprintf("Purity: %.1f%%  |  ARI: %.3f\n", ev_s$purity * 100, ev_s$ari))
```

---

## References

- Ren et al. (2019). Time-series anomaly detection service at Microsoft.
  *KDD 2019*. <https://doi.org/10.1145/3292500.3330680>
- Singh, G., Memoli, F., and Carlsson, G. (2007). Topological methods for
  the analysis of high dimensional data sets and 3D object recognition.
  *Eurographics Symposium on Point-Based Graphics*, 91-100.
- Carlsson, G. (2009). Topology and data.
  *Bulletin of the American Mathematical Society*, 46(2), 255-308.
